RepeatSoaker effect on DEGs detection
========================================================
```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=300, out.width=700, echo=F)
options(replace.assign=TRUE, width=120)
set.seed(1)
```

```{r loadLibraries, warning=FALSE, message=FALSE}
library(biomaRt)
# Use the latest biomart version for mm
mart <- useMart("ensembl", dataset="mmusculus_gene_ensembl")
source("work_scripts//venn5.R")
library(gridExtra)
library(DESeq2)
annotDb <- "org.Mm.eg.db"
library(package=annotDb, character.only=T)
library(GO.db)
library(GOstats)
library(reactome.db)
# Preparing environment for remapping Gene Symbols to Entrez IDs
x <- org.Mm.egSYMBOL2EG
# Get the entrez gene identifiers that are mapped to a gene symbol
mapped_genes <- mappedkeys(x)
# Convert to a list
xx <- as.list(x[mapped_genes])
```

```{r Enrichment, echo=FALSE}
# Perform GO enrichment analysis of topTable produced by limma.
# use use for KEGG of GO enrichment analyses
# Provide which ontology space to use, "MF", "BP", of "CC"
# Returns a list of the results, and total count of significant enrichments
Enrichment <- function(res, use="GO", fileName=NULL, ont="BP")
{
  res.genes <- res
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res.genes]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx) # All Entrez genes
  # Prepare parameters for the enrichment analysis
  if (use == "GO")
    {
    params <- new('GOHyperGParams', geneIds=sel.entrez, universeGeneIds=all.entrez, ontology=ont,
 pvalueCutoff=0.05, conditional=F, testDirection='over', annotation=annotDb)
    }
 else
   {
    params <- new('KEGGHyperGParams', geneIds=sel.entrez, universeGeneIds=all.entrez, pvalueCutoff=0.05, testDirection='over', annotation=annotDb) 
   }
  hgOver <- hyperGTest(params)
  result <- summary(hgOver)
  result <- cbind(result, p.adjust(result$Pvalue, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, nrow(result)))
}
```

```{r GOEnrichment}
# Perform GO enrichment analysis of topTable produced by limma.
# Provide which ontology space to use, "MF", "BP", of "CC"
# Returns a list of the results, and total count of significant enrichments
GOEnrichment <- function(res, ont="BP", fileName=NULL)
{
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx)# All Entrez genes
  # Prepare parameters for the GO enrichment analysis
  params <- new('GOHyperGParams',
                geneIds=sel.entrez,
                universeGeneIds=all.entrez,
                ontology=ont,
                pvalueCutoff=0.05,
                conditional=F,
                testDirection='over',
                annotation="org.Mm.eg.db"
  )
  hgOver <- hyperGTest(params)
  result <- summary(hgOver)
  result <- cbind(result, p.adjust(result$Pvalue, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, nrow(result)))
}
```

```{r reactomeEnrichment, echo=FALSE}
# Perform Reactome pathway enrichment analysis of topTable produced by limma
# Returns a list of the results, and total count of significant enrichments
reactomeEnrichment <- function(res, fileName=NULL)
{
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx); all.entrez <- all.entrez[!is.na(all.entrez)]
  # Prepare pathways and counts
  entrez2path <- as.list(reactomeEXTID2PATHID) # EntrezIDs to pathway IDs mapping
  sel.counts <- table(unlist(entrez2path[sel.entrez])) # Selected gene counts in pathways
  all.counts <- table(unlist(entrez2path[all.entrez])) # All gene counts
  n.sel <- length(sel.entrez) # Total number of selected genes
  n.all <- length(all.entrez) # Total number of all genes
  pathways <- names(sel.counts) # PAthways to test for enrichment
  pathway.names <- as.list(reactomePATHID2NAME) # Pathway ID to name mapping
  # For each pathway, construct 2x2 contingency table and calculate the enrichment p-value using Fisher's exact test
  result <- lapply(pathways, function(p) 
  {
                   sel <- sel.counts[[p]] 
                   all <- all.counts[[p]]
                   pathway.name <- pathway.names[[p]]
                   m <- matrix(c(sel, all, n.sel-sel, n.all-all),
                               byrow=T,nrow=2)
                   p.value <- fisher.test(m, alternative="greater")$p.value
                   data.frame(PathwayID=p, 
                              Pathway=pathway.name,
                              N.Selected=sel,
                              N.In.Category=all,
                              p.value=p.value)
  }
  ) # End lapply
  result <- as.data.frame(do.call(rbind, result)) # Convert list to data frame
  result <- result[order(result$p.value),] # Order by p-value
  result <- cbind(result, p.adjust(result$p.value, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  # result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, sum(result$p.adj < 0.05)))
}
```

```{r biomart}
refseq2gene <- function(x, fileName) {
  # Function to convert NM_ IDs to gene names/descriptions
  genes <- getBM(attributes=c('external_gene_id', 'description'), filters='refseq_mrna', values=x, mart=mart, uniqueRows=T)
  if (!is.null(fileName)) {
    write.table(genes, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(genes)
}
ens2gene <- function(x, fileName=NULL) {
  # Function to convert NM_ IDs to gene names/descriptions
  genes <- getBM(attributes=c('external_gene_id', 'description'), filters='ensembl_gene_id', values=x, mart=mart, uniqueRows=T)
  if (!is.null(fileName)) {
    write.table(genes, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(genes)
}
```

```{r loadData, results='hide'}
readDDS <- function(DIR) {
  sampleFiles <- list.files(path=DIR, pattern=".+txt$")
  sampleFiles <- sampleFiles[grep("control4|mutant1", sampleFiles, invert=T)] # remove CTL4 and MUT1 outliers
  sampleCondition <- sub("^control.*", "control", sampleFiles)
  sampleCondition[sampleCondition != "control"] <- "mutant"
  sampleTable <- data.frame(sampleName = sampleFiles,
                            fileName = sampleFiles,
                            condition = sampleCondition)
  ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                         directory = DIR,
                                         design = ~ condition)
  ddsHTSeq$condition <- factor(ddsHTSeq$condition, levels=c("control", "mutant"))
  return(ddsHTSeq)
}

DIR <- c("data//htseq-count_rev/", "data//htseq-count_rev.75", "data//htseq-count_rev.50", "data//htseq-count_rev.25", "data//htseq-count_rev.01")
ddsHTSEQ <- lapply(DIR, readDDS)
```

Overlap among DEGs identified in the non-soaked data (index [[1]]) and RepeatSoaked data at 75%, 50, 25 and 1% thresholds (subsequent indexes).

```{r diffGeneIDs}
resDDS <- function(DDS, pval=0.01, lfc=0) {
  dds <- DESeq(DDS)
  res <- results(dds)
  res <- res[complete.cases(res), ]
  return(res[res$padj < pval & abs(res$log2FoldChange) > lfc, ])
}

res <- lapply(ddsHTSEQ, resDDS)

plot_venn5(rownames(res[[1]]), rownames(res[[2]]), rownames(res[[3]]), rownames(res[[4]]), rownames(res[[5]]))
```

We observe that soaking diminishes the number of differentially expressed genes. More rigorous soaking (less strict overlap threshold) also increases the number of genes not detected previously.

Withour RepeatSoaker, we have (all genes in the rownames(res[[1]]) oval):

```{r allGenes, results='hide'}
# Get all genes from sig_genes_names[[1]], no repeatSoaker applied
message(paste("Number of differentially expressed PROBES, no RepeatSoaker:", nrow(res[[1]])))
message(paste("Number of differentially expressed GENES, no RepeatSoaker:", length(unique(ens2gene(rownames(res[[1]]))[, 1]))))
```

Some probes map to multiple genes, hence the discrepancy in numbers. 

After trimming the data with all RepeatSoaker settings, we have (the center of the Venn diagram):

```{r commonGenes}
# Get common genes from the center of the Venn diagram
sig_genes_names_common <- Reduce(intersect, list(rownames(res[[1]]), rownames(res[[2]]), rownames(res[[3]]), rownames(res[[4]]), rownames(res[[5]])))
message(paste("Number of differentially expressed PROBES, with RepeatSoaker:", length(sig_genes_names_common)))
message(paste("Number of differentially expressed GENES, with RepeatSoaker:", length(unique(ens2gene(sig_genes_names_common)[, 1]))))
```

Later, we compare these gene lists (Genes without vs. Genes with RepeatSoaker) for biological meaning.

```{r Uniq}
uniq <- function(idx, filename=NULL, raw=F) {
  # Find unique genes for the condition idx
  idxs <- seq(1:length(res)) # All indexes
  # Get unique Ensembl ISa
  uniq.xloc <- setdiff(rownames(res[[idx]]), unique(unlist(sapply(setdiff(idxs, idx), function(x) rownames(res[[x]])))))
  if (raw) {
    return(uniq.xloc)
  } else {
    return(ens2gene(uniq.xloc, filename)) # Vector of NM_ IDs
  }
}
#grid.table(GOEnrichment(as.character(as.matrix(uniq.all[, 2])))[[1]][1:20], )
```

Now, we check what those genes unique to each RepeatSoaker % are. We will look at 3 things:

1) Gene names and their description. Note that not all probes can be mapped to genes, and some probes map to the same gene - therefore, the numbers in the Venn diagram and the tables below differ.

2) GO and Pathway enrichment of those genes, if any. 

```{r GO}
GOKEGG <- function(x, use="GO", filename=NULL) {
  res.go <- Enrichment(x[, 1], use, fileName=filename)
  message("The number of enriched GOs:", res.go[[2]])
  if (res.go[[2]] > 20) {i <- 20} else {i <- res.go[[2]]}
  if (i > 0) {grid.table(res.go[[1]][1:i, ], gp=gpar(fontsize=7))}
}
```

```{r Pathway}
Pathway <- function(x, filename=NULL) {
  res.pathway <- reactomeEnrichment(x[, 1], filename)
  message("The number of enriched pathways:", res.pathway[[2]])
  if (res.pathway[[2]] > 20) {i <- 20} else {i <- res.pathway[[2]]}
  if (i > 0) {grid.table(res.pathway[[1]][1:30, ], gp=gpar(fontsize=7))}
}
```

Genes unique for different RepeatSoaker settings
-------------------------------------------------

Unique genes without RepeatSoaker
---------------------------------

```{r u1}
uniq.all <- uniq(1, "genes_uniq_all.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[, 1]), ], gp=gpar(fontsize=6))  
```

```{r KEGGall}
GOKEGG(uniq.all, use="KEGG", "KEGG_uniq_all.txt")
```

```{r GOall}
GOKEGG(uniq.all, use="GO", "GO_uniq_all.txt")
```

```{r Pathwayall}
Pathway(uniq.all, "PW_uniq_all.txt")
```

Unique genes with 75% RepeatSoaker
----------------------------------

```{r u2}
uniq.r75 <- uniq(2, "genes_uniq_r75.txt") # Condition-specific gene names
grid.table(uniq.r75[order(uniq.r75[,1]), ], gp=gpar(fontsize=6))  
```

```{r KEGGr75}
GOKEGG(uniq.r75, use="KEGG", "KEGG_uniq_r75.txt")
```

```{r GOr75}
GOKEGG(uniq.r75, use="GO", "GO_uniq_r75.txt")
```

```{r Pathwayr75}
Pathway(uniq.r75, "PW_uniq_r75.txt")
```

Unique genes with 50% RepeatSoaker
----------------------------------

```{r u3}
uniq.r50 <- uniq(3, "genes_uniq_r50.txt") # Condition-specific gene names
grid.table(uniq.r50[order(uniq.r50[,1]), ], gp=gpar(fontsize=6))  
```

```{r KEGGr50}
GOKEGG(uniq.r50, use="KEGG", "KEGG_uniq_r50.txt")
```

```{r GOr50}
GOKEGG(uniq.r50, use="GO", "GO_uniq_r50.txt")
```

```{r Pathwayr50}
Pathway(uniq.r50, "PW_uniq_r50.txt")
```

Unique genes with 25% RepeatSoaker
---------------------------------

```{r u4}
uniq.r25 <- uniq(4, "genes_uniq_r25.txt") # Condition-specific gene names
grid.table(uniq.r25[order(uniq.r25[,1]), ], gp=gpar(fontsize=6))  
```

```{r KEGGr25}
GOKEGG(uniq.r25, use="KEGG", "KEGG_uniq_r25.txt")
```

```{r GOr25}
GOKEGG(uniq.r25, use="GO", "GO_uniq_r25.txt")
```

```{r Pathwayr25}
Pathway(uniq.r25, "PW_uniq_r25.txt")
```

Unique genes with 00% RepeatSoaker
---------------------------------

```{r u5}
uniq.r00 <- uniq(5, "genes_uniq_r00.txt") # Condition-specific gene names
grid.table(uniq.r00[order(uniq.r00[,1]), ], gp=gpar(fontsize=6))  
```

```{r KEGGr00}
GOKEGG(uniq.r00, use="KEGG", "KEGG_uniq_r00.txt")
```

```{r GOr00}
GOKEGG(uniq.r00, use="GO", "GO_uniq_r00.txt")
```

```{r Pathwayr00}
Pathway(uniq.r00, "PW_uniq_r00.txt")
```

Other tests
------------

Let's have a look at the distribution of log2 fold change of genes in the leaves of the Venn diagram, as compared with that of the heart of the Venn diagram. It is expected that the leaves may have fold change different from the main DEGs.

```{r boxPlot}
par(oma=c(2, 5, 1, 1))
boxplot(list(all=res[[1]][uniq(1, , raw=T), ]$log2FoldChange,
             r75=res[[2]][uniq(2, , raw=T), ]$log2FoldChange,
             r50=res[[3]][uniq(3, , raw=T), ]$log2FoldChange,
             r25=res[[4]][uniq(4, , raw=T), ]$log2FoldChange,
             r00=res[[5]][uniq(5, , raw=T), ]$log2FoldChange,
             VennHeart=res[[1]][Reduce(intersect, list(rownames(res[[1]]), rownames(res[[2]]), rownames(res[[3]]), rownames(res[[4]]), rownames(res[[5]]))), ]$log2FoldChange),
        ylab="log2FC distribution")
```

We will also check the same for the expression level. The leaves may have higher expression level, hence, more susceptible to the RepeatSoaker.

To be continued.