Understanding the biology of RepeatSoaker effect
=================================================

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=300, out.width=700, echo=F)
options(replace.assign=TRUE, width=120)
set.seed(1)
```

```{r loadLibraries, warning=FALSE, message=FALSE}
annotDb <- "org.Mm.eg.db"
library(package=annotDb, character.only=T)
library(gridExtra)
library(GO.db)
library(GOstats)
library(reactome.db)
# Preparing environment for remapping Gene Symbols to Entrez IDs
x <- org.Mm.egSYMBOL2EG
# Get the entrez gene identifiers that are mapped to a gene symbol
mapped_genes <- mappedkeys(x)
# Convert to a list
xx <- as.list(x[mapped_genes])
```

```{r Enrichment, echo=FALSE}
# Perform GO enrichment analysis of topTable produced by limma.
# use use for KEGG of GO enrichment analyses
# Provide which ontology space to use, "MF", "BP", of "CC"
# Returns a list of the results, and total count of significant enrichments
Enrichment <- function(res, use="GO", fileName=NULL, ont="BP")
{
  res.genes <- res
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res.genes]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx) # All Entrez genes
  # Prepare parameters for the enrichment analysis
  if (use == "GO")
    {
    params <- new('GOHyperGParams', geneIds=sel.entrez, universeGeneIds=all.entrez, ontology=ont,
 pvalueCutoff=0.05, conditional=F, testDirection='over', annotation=annotDb)
    }
 else
   {
    params <- new('KEGGHyperGParams', geneIds=sel.entrez, universeGeneIds=all.entrez, pvalueCutoff=0.05, testDirection='over', annotation=annotDb) 
   }
  hgOver <- hyperGTest(params)
  result <- summary(hgOver)
  result <- cbind(result, p.adjust(result$Pvalue, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, nrow(result)))
}
```


```{r GOEnrichment}
# Perform GO enrichment analysis of topTable produced by limma.
# Provide which ontology space to use, "MF", "BP", of "CC"
# Returns a list of the results, and total count of significant enrichments
GOEnrichment <- function(res, ont="BP", fileName=NULL)
{
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx)# All Entrez genes
  # Prepare parameters for the GO enrichment analysis
  params <- new('GOHyperGParams',
                geneIds=sel.entrez,
                universeGeneIds=all.entrez,
                ontology=ont,
                pvalueCutoff=0.05,
                conditional=F,
                testDirection='over',
                annotation="org.Mm.eg.db"
  )
  hgOver <- hyperGTest(params)
  result <- summary(hgOver)
  result <- cbind(result, p.adjust(result$Pvalue, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, nrow(result)))
}
```

```{r reactomeEnrichment, echo=FALSE}
# Perform Reactome pathway enrichment analysis of topTable produced by limma
# Returns a list of the results, and total count of significant enrichments
reactomeEnrichment <- function(res, fileName=NULL)
{
  # Convert selected and all gene names to Entrez IDs, removing NAs
  sel.entrez <- unlist(xx)[res]; sel.entrez <- sel.entrez[!is.na(sel.entrez)]
  all.entrez <- unlist(xx); all.entrez <- all.entrez[!is.na(all.entrez)]
  # Prepare pathways and counts
  entrez2path <- as.list(reactomeEXTID2PATHID) # EntrezIDs to pathway IDs mapping
  sel.counts <- table(unlist(entrez2path[sel.entrez])) # Selected gene counts in pathways
  all.counts <- table(unlist(entrez2path[all.entrez])) # All gene counts
  n.sel <- length(sel.entrez) # Total number of selected genes
  n.all <- length(all.entrez) # Total number of all genes
  pathways <- names(sel.counts) # PAthways to test for enrichment
  pathway.names <- as.list(reactomePATHID2NAME) # Pathway ID to name mapping
  # For each pathway, construct 2x2 contingency table and calculate the enrichment p-value using Fisher's exact test
  result <- lapply(pathways, function(p) 
  {
                   sel <- sel.counts[[p]] 
                   all <- all.counts[[p]]
                   pathway.name <- pathway.names[[p]]
                   m <- matrix(c(sel, all, n.sel-sel, n.all-all),
                               byrow=T,nrow=2)
                   p.value <- fisher.test(m, alternative="greater")$p.value
                   data.frame(PathwayID=p, 
                              Pathway=pathway.name,
                              N.Selected=sel,
                              N.In.Category=all,
                              p.value=p.value)
  }
  ) # End lapply
  result <- as.data.frame(do.call(rbind, result)) # Convert list to data frame
  result <- result[order(result$p.value),] # Order by p-value
  result <- cbind(result, p.adjust(result$p.value, method="BH")) # Append corrected for multiple testing p-value
  colnames(result)[length(colnames(result))] <- "p.adj"
  # result <- result[result$p.adj < 0.1, ] # Subset the results keeping FDR at 10%
  if (!is.null(fileName)) {
    write.table(result, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(list(result, sum(result$p.adj < 0.05)))
}
```

```{r loadData}
unique <- lapply(c("results//genes_uniq_all.txt", "results//genes_uniq_r75.txt", "results//genes_uniq_r50.txt", "results//genes_uniq_r25.txt", "results//genes_uniq_r00.txt", "results//genes_all.txt", "results//genes_common.txt"), function(x) read.table(x, sep="\t", header=T))
```

```{r GO}
GOKEGG <- function(x, use="GO", filename=NULL) {
  res.go <- Enrichment(x[, 1], use, fileName=filename)
  message("The number of enriched GOs:", res.go[[2]])
  if (res.go[[2]] > 20) {i <- 20} else {i <- res.go[[2]]}
  if (i > 0) {grid.table(res.go[[1]][1:i, ], gp=gpar(fontsize=7))}
}
```

```{r Pathway}
Pathway <- function(x, filename=NULL) {
  res.pathway <- reactomeEnrichment(x[, 1], filename)
  message("The number of enriched pathways:", res.pathway[[2]])
  if (res.pathway[[2]] > 20) {i <- 20} else {i <- res.pathway[[2]]}
  if (i > 0) {grid.table(res.pathway[[1]][1:30, ], gp=gpar(fontsize=7))}
}
```

Our goal is to see whether thrown away genes may be biologically interesting. What are we losing after RepeatSoaker?

We check unique genes from the Venn diagram for GO and pathway enrichment.

Unique genes without RepeatSoaker
---------------------------------

```{r KEGGall}
GOKEGG(unique[[1]], use="KEGG")
```

```{r GOall}
GOKEGG(unique[[1]], use="GO", "GO_uniq_all.txt")
```

```{r Pathwayall}
Pathway(unique[[1]], "PW_uniq_all.txt")
```

Unique genes with 75% RepeatSoaker
----------------------------------

```{r KEGGr75}
GOKEGG(unique[[2]], use="KEGG")
```

```{r GOR75}
GOKEGG(unique[[2]], use="GO", "GO_uniq_r75.txt")
```

```{r PathwayR75}
Pathway(unique[[2]], "PW_uniq_r75.txt")
```

Unique genes with 50% RepeatSoaker
----------------------------------

```{r KEGGr50}
GOKEGG(unique[[3]], use="KEGG")
```

```{r GOR50}
GOKEGG(unique[[3]], use="GO", "GO_uniq_r50.txt")
```

```{r PathwayR50l}
Pathway(unique[[3]], "PW_uniq_r50.txt")
```

Unique genes with 25% RepeatSoaker
---------------------------------

```{r KEGGr25}
GOKEGG(unique[[4]], use="KEGG")
```

```{r GOR25}
GOKEGG(unique[[4]], use="GO", "GO_uniq_r25.txt")
```

```{r PathwayR25}
Pathway(unique[[4]], "PW_uniq_r25.txt")
```

Unique genes with 00% RepeatSoaker
---------------------------------

```{r KEGGr00}
GOKEGG(unique[[5]], use="KEGG")
```

```{r GOR00}
GOKEGG(unique[[5]], use="GO", "GO_uniq_r00.txt")
```

```{r PathwayR00}
Pathway(unique[[5]], "PW_uniq_r00.txt")
```

Comparing genes without and with RepeatSoaker treatment
========================================================
Genes without RepeatSoaker
----------------------------
```{r KEGG_noRS}
GOKEGG(unique[[6]], use="KEGG")
```

```{r GO_noRS}
GOKEGG(unique[[6]], use="GO", "GO_genes_all.txt")
```

```{r Pathway_noRS}
Pathway(unique[[6]], "PW_genes_all.txt")
```

Genes with RepeatSoaker
--------------------------
```{r KEGG_wRS}
GOKEGG(unique[[7]], use="KEGG", "KEGG_genes_common.txt")
```

```{r GO_wRS}
GOKEGG(unique[[7]], use="GO", "GO_genes_common.txt")
```

```{r Pathway_wRS}
Pathway(unique[[7]], "PW_genes_common.txt")
```

