RepeatSoaker effect on DEGs detection
========================================================
```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=T, dpi=300, out.width=700, echo=F)
options(replace.assign=TRUE, width=120)
set.seed(1)
```

```{r loadLibraries, warning=FALSE, message=FALSE}
library(biomaRt)
# Use the latest biomart version for mm
mart <- useMart("ensembl", dataset="mmusculus_gene_ensembl")
library("cummeRbund")
source("work_scripts//venn5.R")
library(gridExtra)
```

```{r biomart}
refseq2gene <- function(x, fileName) {
  # Function to convert NM_ IDs to gene names/descriptions
  genes <- getBM(attributes=c('external_gene_id', 'description'), filters='refseq_mrna', values=x, mart=mart, uniqueRows=T)
  if (!is.null(fileName)) {
    write.table(genes, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(genes)
}
```

```{r loadData}
cuff_data <- lapply(list(all="data//cuffdiff-/", r75="data//cuffdiff-.75/", r50="data//cuffdiff-.50/", r25="data//cuffdiff-.25", r01="data//cuffdiff-.00/"), readCufflinks) # , r00="data//cuffdiff-.00", rebuild=TRUE
```

Overlap among DEGs identified in the non-soaked data (index [[1]]) and RepeatSoaked data at 75%, 50, 25 and 1% thresholds (subsequent indexes).

```{r diffGeneIDs}
DEGcount <- function(cuff_data) {
  # Function to get tables of cuffdiff analysis
  gene_diff_data <- diffData(genes(cuff_data),"CTRL","MUT",features=T)
  sig_gene_data <- subset(gene_diff_data, (significant == 'yes')) # Get significant only
  return(sig_gene_data)
}
# Get RepeatSoaker-specific results in a list
sig_genes <- lapply(cuff_data, DEGcount)
## 5-way venn diagram
# Get condition-specific lists of gene_id
sig_genes_names <- lapply(sig_genes, function(x) x$gene_id)
sapply(sig_genes_names, length)
names(sig_genes_names) <- names(cuff_data)
#do.call(plot_venn5, sig_genes_names)
plot_venn5(sig_genes_names[[1]], sig_genes_names[[2]], sig_genes_names[[3]], sig_genes_names[[4]], sig_genes_names[[5]])
```

We observe that soaking diminishes the number of differentially expressed genes. More rigorous soaking (less strict overlap threshold) also increases the number of genes not detected previously.

Withour RepeatSoaker, we have (all genes in the sig_genes_names[[1]] oval):

```{r allGenes, results='hide'}
# Get all genes from sig_genes_names[[1]], no repeatSoaker applied
sig_genes_nmids_all <- as.character(featureNames(getGenes(cuff_data[[1]], sig_genes_names[[1]]))[, 2])
message(paste("Number of differentially expressed PROBES, no RepeatSoaker:", length(sig_genes_nmids_all)))
sig_genes_nmids_all <- unique(unlist(lapply(strsplit(sig_genes_nmids_all, ","), function(v) {append(vector(), unique(v))})))
message(paste("Number of differentially expressed GENES, no RepeatSoaker:", length(sig_genes_nmids_all)))
tmp <- refseq2gene(sig_genes_nmids_all, "genes_all.txt")
```

Some probes map to multiple genes, hence the discrepancy in numbers. 

After trimming the data with all RepeatSoaker settings, we have (the center of the Venn diagram):

```{r commonGenes}
# Get common genes from the center of the Venn diagram
sig_genes_names_common <- Reduce(intersect, list(sig_genes_names[[1]], sig_genes_names[[2]], sig_genes_names[[3]], sig_genes_names[[4]], sig_genes_names[[5]]))
message(paste("Number of differentially expressed PROBES, with RepeatSoaker:", length(sig_genes_names_common)))
sig_genes_nmids_common <- as.character(featureNames(getGenes(cuff_data[[1]], sig_genes_names_common))[, 2])
sig_genes_nmids <- unique(unlist(lapply(strsplit(sig_genes_nmids_common, ","), function(v) {append(vector(), unique(v))})))
message(paste("Number of differentially expressed GENES, with RepeatSoaker:", length(sig_genes_nmids)))
tmp <- refseq2gene(sig_genes_nmids, "genes_common.txt")
```

Later, we compare these gene lists (Genes without vs. Genes with RepeatSoaker) for biological meaning.

```{r Uniq}
uniq <- function(idx) {
  # Find unique genes for the condition idx
  idxs <- seq(1:length(cuff_data)) # All indexes
  # Get unique XLOC IDs
  uniq.xloc <- setdiff(sig_genes_names[[idx]], unique(unlist(sapply(setdiff(idxs, idx), function(x) sig_genes_names[[x]]))))
  # Convert XLOC IDs to NM_ IDs. Some map to multiple NM_ IDs, comma separated
  uniq.nmid <- as.character(featureNames(getGenes(cuff_data[[idx]], uniq.xloc))[, 2])
  # splic comma-separated IDs and append them to a vector of unique NM_ IDs
  uniq.nmid <- unique(unlist(lapply(strsplit(uniq.nmid, ","), function(v) {append(vector(), unique(v))})))
  return(uniq.nmid) # Vector of NM_ IDs
}
#grid.table(GOEnrichment(as.character(as.matrix(uniq.all[, 2])))[[1]][1:20], )
```

Now, we check what those genes unique to each RepeatSoaker % are. We will look at 3 things:

1) Gene names and their description. Note that not all probes can be mapped to genes, and some probes map to the same gene - therefore, the numbers in the Venn diagram and the tables below differ.

2) GO and Pathway enrichment of those genes, if any. (featureNames from cummeRbund conflicts with org.Mm.ed.db package, see **Analysis_RepeatSoaker_genes.md** document)

Genes unique for different RepeatSoaker settings
-------------------------------------------------

```{r u1}
message("Unique genes for non-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(1), "genes_uniq_all.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

```{r u2}
message("Unique genes for 75%-RepeatSoaked data")
uniq.r75 <- refseq2gene(uniq(2), "genes_uniq_r75.txt") # Condition-specific gene names
grid.table(uniq.r75[order(uniq.r75[,1]), ], gp=gpar(fontsize=6))  
```

```{r u3}
message("Unique genes for 50%-RepeatSoaked data")
uniq.r50 <- refseq2gene(uniq(3), "genes_uniq_r50.txt") # Condition-specific gene names
grid.table(uniq.r50[order(uniq.r50[,1]), ], gp=gpar(fontsize=6))  
```

```{r u4}
message("Unique genes for 25%-RepeatSoaked data")
uniq.r25 <- refseq2gene(uniq(4), "genes_uniq_r25.txt") # Condition-specific gene names
grid.table(uniq.r25[order(uniq.r25[,1]), ], gp=gpar(fontsize=6))  
```

```{r u5}
message("Unique genes for 0%-RepeatSoaked data")
uniq.r00 <- refseq2gene(uniq(5), "genes_uniq_r00.txt") # Condition-specific gene names
grid.table(uniq.r00[order(uniq.r00[,1]), ], gp=gpar(fontsize=6))  
```


```{r plotSD}
plotSumStats <- function(idx){
  mtx1 <- repFpkmMatrix(genes(cuff_data[[idx]]))
  mtx1.x <- seq(1:nrow(mtx1))
  bpts <- pretty(mtx1.x, n=100)
  INDEX <- cut(mtx1.x, bpts, include.lowest=T)
  #mtx1.y <- tapply(log2(1 + sort(rowMeans(mtx1))), INDEX, sd)
  mtx1.y <- tapply(log2(apply(mtx1[order(rowMeans(mtx1)), ], 1, sd)), INDEX, mean)
  mtx1.x <- seq(1:(length(bpts) - 1))
  if (idx ==1) plot(mtx1.x, mtx1.y, type="n") # Create empty plot onl once
  lines(mtx1.x, mtx1.y, type="l", lwd=3, lty=linetype[idx], col=colors[idx], pch=plotchar[idx]) # Then, overlay lines
}
colors <-rainbow(length(cuff_data))
linetype <- c(1:length(cuff_data))
plotchar <- seq(16, 16 + length(cuff_data), 1)
sapply(seq(1:length(cuff_data)), function(i) plotSumStats(i))
legend(0, 10, names(cuff_data), cex=0.8, col=colors, pch=plotchar, lty=linetype, title="condition")

```

