RepeatSoaker effect on DEGs detection
========================================================
```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=300, out.width=700, echo=F)
options(replace.assign=TRUE, width=120)
set.seed(1)
```

```{r loadLibraries, warning=FALSE, message=FALSE}
library(biomaRt)
# Use the latest biomart version for mm
mart <- useMart("ensembl", dataset="mmusculus_gene_ensembl")
library("cummeRbund")
source("work_scripts//venn5.R")
library(gridExtra)
```

```{r biomart}
refseq2gene <- function(x, fileName) {
  # Function to convert NM_ IDs to gene names/descriptions
  genes <- getBM(attributes=c('external_gene_id', 'description'), filters='refseq_mrna', values=x, mart=mart, uniqueRows=T)
  if (!is.null(fileName)) {
    write.table(genes, paste("results//", fileName, sep=""), sep="\t", row.names=F)
  }
  return(genes)
}
```

```{r loadData}
cuff_data <- lapply(list(all="data//cuffdiff-/", r75="data//cuffdiff-.75/", r50="data//cuffdiff-.50/", r25="data//cuffdiff-.25", r01="data//cuffdiff-.01/"), readCufflinks) # , r00="data//cuffdiff-.00", rebuild=TRUE
```

Overlap among DEGs identified in the non-soaked data (index [[1]]) and RepeatSoaked data at 75%, 50, 25 and 1% thresholds (subsequent indexes).

```{r diffGeneIDs}
DEGcount <- function(cuff_data) {
  # Function to get tables of cuffdiff analysis
  gene_diff_data <- diffData(genes(cuff_data),"CTRL","MUT",features=T)
  sig_gene_data <- subset(gene_diff_data, (significant == 'yes')) # Get significant only
  return(sig_gene_data)
}
# Get RepeatSoaker-specific results in a list
sig_genes <- lapply(cuff_data, DEGcount)
## 5-way venn diagram
# Get condition-specific lists of gene_id
sig_genes_names <- lapply(sig_genes, function(x) x$gene_id)
sapply(sig_genes_names, length)
names(sig_genes_names) <- names(cuff_data)
#do.call(plot_venn5, sig_genes_names)
plot_venn5(sig_genes_names[[1]], sig_genes_names[[2]], sig_genes_names[[3]], sig_genes_names[[4]], sig_genes_names[[5]])

```

We observe that soaking diminishes the number of differentially expressed genes. More rigorous soaking (less strict overlap threshold) also increases the number of genes not detected previously.

```{r Uniq}
uniq <- function(idx) {
  # Find unique genes for the condition idx
  idxs <- seq(1:length(cuff_data)) # All indexes
  # Get unique XLOC IDs
  uniq.xloc <- setdiff(sig_genes_names[[idx]], unique(unlist(sapply(setdiff(idxs, idx), function(x) sig_genes_names[[x]]))))
  # Convert XLOC IDs to NM_ IDs. Some map to multiple NM_ IDs, comma separated
  uniq.nmid <- as.character(featureNames(getGenes(cuff_data[[idx]], uniq.xloc))[, 2])
  # splic comma-separated IDs and append them to a vector of unique NM_ IDs
  uniq.nmid <- unique(unlist(lapply(strsplit(uniq.nmid, ","), function(v) {append(vector(), unique(v))})))
  return(uniq.nmid) # Vector of NM_ IDs
}
#grid.table(GOEnrichment(as.character(as.matrix(uniq.all[, 2])))[[1]][1:20], )
```

Now, we check what those genes unique to each RepeatSoaker % are. We will look at 3 things:

1) Gene names and their description. Note that not all probes can be mapped to genes, and some probes map to the same gene - therefore, the numbers in the Venn diagram and the tables below differ.

2) GO enrichment of those genes, if any. (featureNames from cummeRbund conflicts with org.Mm.ed.db package, currently not functional)

3) Reactome pathway enrichment of those genes, if any. (TODO)

The goal is to (subjectively) judge what we are loosing. 

```{r u1}
message("Unique genes for non-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(1), "genes_uniq_all.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

```{r u2}
message("Unique genes for 75%-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(2), "genes_uniq_r75.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

```{r u3}
message("Unique genes for 50%-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(3), "genes_uniq_r50.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

```{r u4}
message("Unique genes for 25%-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(4), "genes_uniq_r25.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

```{r u5}
message("Unique genes for 1%-RepeatSoaked data")
uniq.all <- refseq2gene(uniq(5), "genes_uniq_r01.txt") # Condition-specific gene names
grid.table(uniq.all[order(uniq.all[,1]), ], gp=gpar(fontsize=6))  
```

TODO: What we are gaining.

